Hypoxia uncouples HIF target gene transcription and glycolytic flux in proliferating primary cells

### 2021-05-10 inst/manuscript/manuscript.Rmd
revisions to rough draft

---
title: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"

author:
- Courtney A. Copeland: 
    institute: 1
- Benjamin A. Olenchock: 
    institute: 1, 2
- David R. Ziehr:
    institute: 1, 4
- Sarah McGarrity: 
    institute: 1, 5
- Jamey D. Young: 
    institute: 3
- Joseph Loscalzo: 
    institute: 1
- William M. Oldham: 
    institute: 1
    email: woldham@bwh.harvard.edu
    correspondence: true
institute:
- 1: Department of Medicine, Brigham and Women's Hospital and Harvard Medical School, Boston, MA
- 2: Regeneron Pharmaceuticals, Tarrytown, NY
- 3: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN
- 4: Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, MA
- 5: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland

bibliography_main: library.bib
bibliography_software: packages.bib
csl: cell-metabolism.csl
link-citations: yes

output: 
  bookdown::word_document2:
    reference_docx: template.docx
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      - '--lua-filter=multiple-bibliographies.lua'

compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
header-includes:
- \usepackage{lineno}
- \linenumbers
- \setlength\linenumbersep{1cm}
- \usepackage{setspace}
- \doublespacing
- \makeatletter\renewcommand*{\fps@figure}{htbp}\makeatother
- \setlength{\parskip}{0.75\baselineskip plus 2pt minus 2pt}
- \usepackage{float}
---

@@

knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  out.width = "50%", 
  fig.align = "center", 
  fig.pos = "H", 
  out.extra = ""
)

@@

withr::with_dir(here::here(), {
  biomass <- targets::tar_read(biomass)
  resources <- targets::tar_read(resources_table)
})

@@

# Summary 
Hypoxia is an important environmental stimulus that causes transcriptional and metabolic reprogramming in cells to facilitate their survival. Here, we performed stable isotope tracing and metabolic flux analyses of proliferating primary cells in hypoxia. Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells in our models. No evidence for increased glutaminolysis or reductive carboxylation was observed. While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect. Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, reveals distinct molecular responses to to hypoxia and pharmacologic prolyl hydroxylase inhibition. Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other regulatory factors override the effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.

@@

Metazoan cells depend on aerobic respiration to meet cellular energy demands. With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980, @RN1198]. Overall, these changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain energy production and minimize the formation of reactive oxygen species [@RN2525].

While this "glycolytic shift" of primary carbon metabolism is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953, @RN1603, @RN1198]. Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395, @RN1447, @RN1438, @RN634, @RN1200, @RN975]. For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634]. By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated in these cells. This information would provide important context for understanding the extent to which cancer cell metabolism responds differently to hypoxic stress. Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.

To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions. We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α activation by the prolyl hydroxylase (PHD) inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state. Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (**Figure \@ref(fig:f1)D-H**). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3**). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Supplementary Figure 1**). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program. Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; **Supplementary Figure 3**). In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628]. Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (Supplementary Figure 3), suggesting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f2)**). Similar to hypoxia, BAY decreased cell growth rate (**Figure \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figure \@ref(fig:f2)C-G**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f2)H**). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f2)J**).

## Stable isotope incorporation suggests preservation of the intracellular metabolic program in hypoxia

To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figure \@ref(fig:f3), Supplementary Figures 4-5**). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figure \@ref(fig:f3)A-B**).

Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater). This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells. BAY treatment recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions. Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY. These findings were similar in PASMCs (**Supplementary Figure 5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions. Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis. The overall fraction of M5-citrate in these cells was low (< 6%).

@@

To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure \@ref(fig:f3)A, Supplementary Figure 6**). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figure \@ref(fig:f4), Supplementary Figure 6, Supplementary Tables 1-3**). Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis. Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Supplementary Figure 6C**). In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (**Figure \@ref(fig:f4)A**) with a significant increase in pentose phosphate pathway flux. Similar findings were observed in PASMC flux maps (**Supplementary Figure 6, Supplementary Table 3**).

Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of a decreased growth rate. After normalizing metabolite fluxes in normoxia and hypoxia to the associated growth rate, modest increases (~10%) in glycolytic flux were observed (Supplementary Figure 4D). This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program. Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.

@@

Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1-3). First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment and by hypoxia treatment in PASMCs.

Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs exposed to hypoxia. Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment. This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366]. To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figure \@ref(fig:f5), Supplementary Figure 5**). Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figure \@ref(fig:f5), Supplementary Figure 4**). Interestingly, a similar increase in labeling was not observed in PASMCs (Supplementary Figure 5).

Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high. Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*. In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by `[`^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization. Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718]. In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG). This is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668]. Together, these data suggest that lactate makes a modest (~5% carbon) contribution to this process.

@@

To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figure \@ref(fig:f6)A-C**). LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen. Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen. Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux. These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.

To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figure \@ref(fig:f6)D-F**). Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f6)D**). These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition. Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (**Figure \@ref(fig:f6)E**). Several patterns emerged from this analysis (**Figure \@ref(fig:f6)F**). Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in these proliferating LFs [@RN1603]. The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine. Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA). A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the tricarboxlyic acid cycle to be the most enriched KEGG metabolite set. Leading edge analysis demonstrates the negative enrichment scores associated with all of the TCA metabolites detected by our platform. This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models were hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f4)**).

In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state. As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figure \@ref(fig:f6)I-K**). As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^. Interestingly, while BAY treatment increased the levels of NADH, a concomitant increase in NAD^+^ resulted in preservation of the NADH/NAD^+^ ratio. As NADH accumulation is a putative inhibitor of glycolytic flux [@RN2210], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.

@@

>>>>

---
title: "Hypoxia uncouples HIF target gene transcription and glycolytic flux in proliferating primary cells"

author:
- Courtney A. Copeland: 
    institute: 1
- Benjamin A. Olenchock: 
    institute: 1, 2
- David R. Ziehr:
    institute: 1, 4
- Sarah McGarrity: 
    institute: 1, 5
- Kevin Leahy: 
    institute: 1
- Jamey D. Young: 
    institute: 3
- Joseph Loscalzo: 
    institute: 1
- William M. Oldham: 
    institute: 1
    email: woldham@bwh.harvard.edu
    correspondence: true
institute:
- 1: Department of Medicine, Brigham and Women's Hospital and Harvard Medical School, Boston, MA
- 2: Regeneron Pharmaceuticals, Tarrytown, NY
- 3: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN
- 4: Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, MA
- 5: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland

bibliography: 
- library.bib
- packages.bib
csl: cell-metabolism.csl
link-citations: yes

output: 
  bookdown::word_document2:
    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      # - '--lua-filter=multiple-bibliographies.lua'

compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
---

@@

knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  out.width = "50%", 
  out.extra = ""
)

@@

withr::with_dir(here::here(), {
  biomass <- targets::tar_read(biomass)
  resources <- targets::tar_read(resources_table)
  figs <- targets::tar_load(targets::matches("m\\d_figure"))
})

@@

# Summary 
Hypoxia is an important environmental stimulus that causes transcriptional and metabolic reprogramming in cells to facilitate their survival. Here, we performed stable isotope tracing and metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia. Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells. While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment. Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition. Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.

@@

Metazoan cells depend on aerobic respiration to meet cellular energy demands. With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198]. Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes in normoxia and targeted for proteasomal degradation. Prolyl hydroxlyase enzymes (PHDs) are the principal oxygen sensors in metazoan cells [@RN999]. PHDs are α-ketoglutrate dioxygenase enzymes that require molecular oxygen to catalyze this protein modification. When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and its associated transcriptional program. Overall, the changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and minimize the formation of reactive oxygen species (ROS) [@RN2525].

While a "glycolytic shift" of primary carbon metabolism in hypoxia is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953; @RN1603; @RN1198]. Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395; @RN1447; @RN1438; @RN634; @RN1200; @RN975]. For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634]. By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of reductive carboxylation and aspartate biosynthesis remain to be elucidated in these cells. This information would provide important context for understanding fundamental differences in how cancer cell metabolism responds to hypoxic stress. Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.

To test this hypothesis, we developed models of bioenergetic carbon flux in human primary lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured under 21% or 0.5% oxygen conditions. We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1 transcriptional program. In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells.

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state. LFs cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)D-H**). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:f1)I-J, S1-S3**). Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program. Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and in PASMCs (**Figure S3**). In addition to lactate, the absolute fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628]. Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (**Figure S3**), suggesting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f2)**). Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:f2)C-G**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f2)H**). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f2)J**).

## Stable isotope labeling suggests preservation of the intracellular metabolic architecture in hypoxia

To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figures \@ref(fig:f3), S4-S5**). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:f3)A-B**).

Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater). This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells. BAY treatment recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions. Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY. These findings were similar in PASMCs (**Figure S5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions. Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis. The overall fraction of M5-citrate in these cells was low (< 6%).

@@

To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:f3)A, S6**). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:f4), S6, Tables S1-S3**). Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis. Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S6C**). In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (**Figure \@ref(fig:f4)A**) with a significant increase in pentose phosphate pathway flux. Similar findings were observed in PASMC flux models (**Figure S6, Table S3**).

Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of decreased growth rate. After normalizing metabolite fluxes in normoxia and hypoxia to the  growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S4D**). This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program. Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.

@@

Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (**Tables S1-S3**). First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia in LFs, is modestly increased by BAY treatment, and is increased by hypoxia in PASMCs.

Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs exposed to hypoxia. Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment. This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366]. To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:f5), S5**). Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figures \@ref(fig:f5), S4**). Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, these labels were not incorporated into the TCA cycle as observed in LFs (**Figure S5**).

Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high. Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*. In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization. Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718]. In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG). This is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668]. Together, these data suggest that lactate makes a modest (~5% carbon) contribution to this process.

@@

To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:f6)A-C**). LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen. Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen. Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux. These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.

To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:f6)D-F**). Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f6)D**). These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition. Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (**Figure \@ref(fig:f6)E**). Several patterns emerged from this analysis (**Figure \@ref(fig:f6)F**). Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs [@RN1603]. The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine. Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA). A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the TCA cycle to be the most enriched KEGG metabolite set. Leading edge analysis demonstrates the negative enrichment scores associated with all of the TCA metabolites detected by our platform. This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models were hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f4)**).

In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state. As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:f6)I-K**). As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^. Interestingly, while BAY treatment increased the levels of NADH, a concomitant increase in NAD^+^ resulted in preservation of the NADH/NAD^+^ ratio. As NADH accumulation is a putative inhibitor of glycolytic flux [@RN2210], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.

@@

Table S1: **Lung fibroblast fluxes in 21% and 0.5% oxygen.**

Table S2: **Lung fibroblast fluxes following DMSO and BAY treatment.**
